---
title: "Analysis of AI ML Data Sciences Job Market And Salary Trends(USA)"
author: "Chinmayee Raman"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output:
  html_document:
    fig_width: 8
    code_folding: hide
    theme: journal
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=9, fig.height=6,
              echo=FALSE, warning=FALSE, message=FALSE)
library(plotly)
library(tidyverse)
library(rmarkdown)
library(sampling)
options(warn=-1)
```

# Dataset Details

This data is compiled from anonymous salary surveys filled out by working
professionals in the AI, ML, and data science fields, and published under the
CC0 license. This covers data collected from the years 2020 to 2024. This file
contains structured salary information across different roles including
experience levels, job title, employment types, and geographical locations.

As we can see from the pie chart, an overwhelming majority(88.9%) of the survey
respondents are working for companies in the USA.Thus, the data set was filtered
for the same and all analysis in this project is for companies located in the
USA.

```{r echo=TRUE}
#read data from csv files into tibble for global_ml_ai_salaries
# read data from excel into tibble for file country iso code
################################################################

              
inpFile1 = "global_ai_ml_data_salaries.csv"
globalSal = as_tibble(read.csv(inpFile1))
plot_ly(labels = globalSal$company_location,type = "pie",width =500,
        height = 400) %>%
  layout(title = "Jobs by Country")
#filter for company locations in the USA
dsSal= filter(globalSal, company_location == "US")
paged_table(dsSal)

```

## Column Details

The column information relevant to our analysis is shown below:

1.  work_year: The year the salary was paid
2.  experience_level: The experience level in the job during the year
    -   (EN: Entry-level/Junior, MI: Mid-level/Intermediate, SE: Senior-level/
        Expert, EX: Executive-level/Director)
3.  job_title: The role worked in during the year.
4.  salary_in_usd: The salary in USD
5.  remote_ratio: The overall amount of work done remotely
    -   0: No remote work, 50: Partially remote/hybrid, 100: Fully remote
6.  company_location: data filter for company_location = "US"
7.  company_size: The average number of people that worked for the company
    during the year
    -   S: less than 50 employees, M: 50 to 250 employees, L: more than 250
        employees

# Objective

Based on the data for the AI ML Data Science the objective of the project is the
following:

1.  Analyze the salary distribution in the AI ML Data Science Sector in the USA.
2.  Prove applicability of the Central Limit Theorem
3.  Analyze applicability of sampling methods to determine population parameters
4.  What size companies are hiring more in this data science sector
5.  What are the top 10 job titles in the industry by frequency
6.  Analyze trends in salary distribution of top 10 job titles
7.  What is the average salary by top 10 job titles, and experience level
8.  What has been the average salary growth in this sector over the past 4 years
    by top 10 job titles
9.  What are the top 10 job titles with the highest average salaries
10. How is the salary distributed by experience across the top 10 job titles
    with highest average salaries
11. Are salaries in this sector directly correlated to job experience
12. What has been the trend in average salaries over the work years
13. Analyze the remote work trends by year

# Salary Distribution

Shown below is the histogram and the box plot for the Salary distribution.

```{r echo=TRUE}

# Plotting histogram for Salary Data
x <- dsSal$salary_in_usd
m = median(x)
mu = round(mean(x),2)
sd = sd(x)

heading = "ML AI Salary Distribution(US Companies)"
#pTitle= list(text=heading, y = 0.98, x = 0.5, xanchor = 'center', yanchor =  'top')
pTitle= list(text=heading)
xTitle = list(title = "Salary in USD")
yTitle = list(title = "Count")
                
plot_ly( x = ~x,type = "histogram",name = "Population Distribution")  %>%
layout(title = pTitle,
       xaxis = xTitle, 
       yaxis = yTitle ,
       legend = list(x = .6,y=.9)) %>% 
  add_lines(x = c(m,m),y = c(0,900),type = "scatter",mode = 'lines'
            ,name = paste("Median Salary in USD =",m) )|>
  add_trace(x = c(mu,mu),y = c(0,900),type = "scatter",mode = 'lines'
            ,name = paste("Mean Salary in USD =",mu))

f = fivenum(x)
IQR = f[4] -f[2]
lOutLimit = f[2] - 1.5*IQR
uOutLimit = f[4] + 1.5*IQR
t = "Boxplot of Salary Distribution(US Companies)"
t = paste(t,"<br>Lower Fence", lOutLimit,"Min:",f[1] ,"Q1:",f[2],"Median:",f[3],
"Q3:",f[4],"Max:",f[5], "Upper Fence", uOutLimit)

plot_ly(x = ~x, type = "box") %>%
  layout( title = list(text= t,font = list(size =14,color ="darkblue",face= "bold")), 
          xaxis = list (title = "Salary in USD",nticks =20))

y <- dsSal$salary_in_usd > 319900
cat (" The number of outliers exceeding the Upper Fence(Upper Outer Limit) of:",uOutLimit,"USD are:" , length(y[y==TRUE]),"\n","The Standard Deviation of the Salary Distribution is:", sd(dsSal$salary_in_usd),"USD","\n","The Mean is",
     mean(dsSal$salary_in_usd),"USD","\n","The Five Number Summary in USD is:",t) 
```

## Findings and Inference

1.  The Salary follows a normal distribution and is skewed to the right,
    indicating outliers to the right of the distribution.

2.  There are 328 people whose salaries are higher than the upper outlier limit
    of \$319,900

3.  The middle 50% of the salaries range from \$110,900 to \$194,500

4.  The five number summary is:

    ```         
    Lower Fence -14500 Min: 20000 Q1: 110900 Median: 148500 Q3: 194500 Max: 750000 Upper Fence 319900
    ```

5.  The standard deviation is: \$66289.21 and mean is: \$157799.72

# Central Limit Theorem

The Applicability of Central Theorem is shown by using the salary attribute of
the data set. Shown below are the histograms of sample means of 1000 random
samples of sample size 10, 20 , 30, and 50 respectively.

```{r echo=TRUE}
# Central Limit Theorem
# draw 1000 samples of size 10 , 20 , 30 , 50 from dsSal$salary_in_usd

#set seed 
set.seed(6603)
# Set number of samples
samples <- 1000
#sample size = 10
size <- 10
xbar <- numeric(samples)
for (i in 1: samples) {
    xbar[i] <- mean(sample(dsSal$salary_in_usd, size, replace = FALSE))
}
y1 <- xbar
s1 <- size
# Plotting histogram for Earnings Data of sample size 10
fig1 <- plot_ly( x = ~y1,   type = "histogram", name = "Sample Size: 10")
#sample size = 20
size <- 20
xbar <- numeric(samples)
for (i in 1: samples) {
    xbar[i] <- mean(sample(dsSal$salary_in_usd, size, replace = FALSE))
}
y2  <- xbar
s2 <- size
# cat("Sample Size =", size, "Mean =", mean(xbar)," SD =", sd(xbar), 
#     "Theoretical SD =", sd/sqrt(size), "\n")
fig2 <- plot_ly( x = ~y2, type = "histogram",name = "Sample Size: 20")
#sample size = 30
size <- 30
xbar <- numeric(samples)
for (i in 1: samples) {
    xbar[i] <- mean(sample(dsSal$salary_in_usd, size, replace = FALSE))
}
y3 <-  xbar
s3 <- size

# Plotting histogram for Earnings Data of sample size 30
fig3 <- plot_ly( x = ~y3,type = "histogram",name = "Sample Size: 30")

#sample size = 50
size <- 50
xbar <- numeric(samples)
for (i in 1: samples) {
    xbar[i] <- mean(sample(dsSal$salary_in_usd, size, replace = FALSE))
}
y4 <- xbar
s4 <- size
# cat("Sample Size =", size, "Mean =", mean(xbar)," SD =", sd(xbar), 
#     "Theoretical SD =", sd/sqrt(size), "\n")
#Plotting histogram for Earnings Data of sample size 50
fig4 <- plot_ly( x = ~y4, type = "histogram",  name = "Sample Size: 50"  ) 
fig4 <- subplot(fig1,fig2,fig3,fig4,nrows = 4,shareX = T) 
fig4 <- fig4 %>%  layout(title = "Histogram of Sample Means of Salary",
         xaxis = list(title = "Sample Means of Salary(USD)"))
fig4
```

```{r echo = TRUE}
#printing out the means and sigmas
sd = sd(dsSal$salary_in_usd)
mu = mean(dsSal$salary_in_usd)
cat(" Population Mean =", mu,   "Population SD =", sd, "\n",
    "Sample Size =", s1, "Mean =", mean(y1)," SD =", sd(y1), 
    "Theoretical SD =", sd/sqrt(10), "\n",
    "Sample Size =", s2, "Mean =", mean(y2)," SD =", sd(y2), 
    "Theoretical SD =", sd/sqrt(20), "\n",
     "Sample Size =", s3, "Mean =", mean(y3)," SD =", sd(y3), 
    "Theoretical SD =", sd/sqrt(30), "\n",
    "Sample Size =", s4, "Mean =", mean(y4)," SD =", sd(y4), 
    "Theoretical SD =", sd/sqrt(50), "\n")
```

## Findings and Inference

1.  From the results shown above we can see that the sample means are close to
    the population mean of \$157799.7.The sample mean of \$157615.1 for sample
    size = 50 is closest to the population mean of \$157799.7. The sample means
    for sample size = 10, 20, and 30 are within \$1000 of the population mean.
2.  As the sample size increases from 10 to 50 we can see that the standard
    deviation decreases.
3.  The values of the actual standard deviations for all the sample sizes are
    close equal to the theoretical standard deviation as shown above.

# Sampling Methods

The applicability of the various sampling methods was analyzed by comparing the
salary distribution of the population versus the salary distribution the
sampling methods. The following sampling methods were used for sample size n =
100.

1.  Simple Random Sampling with Replacement
2.  Systematic Sampling (Unequal Probabilities)\
    The sampling was done after finding the proportion of salaries in the
    population and picking the samples in the same proportion.
3.  Systematic Sampling (Equal Probabilities)\
    Here we divided the data into 100 groups and picked the sample data from the
    same position r (r is determined randomly) from each group. Since we used
    the ceiling function to split the data into groups, the sample value taken
    from the last group was = NA (there was no data in the r th position in this
    group). This value was removed from the sample. Hence the sample size
    reduced to 99.
4.  Stratified Sampling\
    The data was divided into strata based on the company size (small, medium
    and large). Due to rounding error while getting the proportion, there the
    was no sample picked for company size = small. Hence a sample was added for
    company size = small, and the number of samples increased to 101.

```{r echo = TRUE}
library(sampling)
#plotting histogram of salary data for US
# Plotting histogram for Salary Data
x <- dsSal$salary_in_usd
m = median(x)
mu = round(mean(x),2)
sd = sd(dsSal$salary_in_usd)

p = plot_ly( x = ~x,type = "histogram",name = "Population Distribution",
             histnorm = "probability",nbinsx = 25)  %>%
   add_trace(x = c(mu,mu),y = c(0,0.4),type = "scatter",mode = 'lines'
            ,name = paste("Mean Salary in USD =",mu))

#set sample size = 50
n = 100
#set 
N = nrow(dsSal)
#get simple random sampling with replacement for sample  size = 100 
# set seed
set.seed(6000)
s <- srswr(n,N )
#s[s != 0]

#getting the selected rows with reps
rows <- (1:N)[s!=0]
rows <- rep(rows, s[s != 0])
sample.1 <- dsSal[rows, ]
x1<- sample.1$salary_in_usd
m = median(x1)
mu = round(mean(x1),2)
sd = sd(x1)

heading = "Simple Random Sampling(With Replacement)"
p1=plot_ly(x = ~x1,type = "histogram",name = "Simple Random Sampling With Replacement",
         histnorm = "probability",nbinsx = 15)  %>%
  add_trace(x = c(mu,mu),y = c(0,0.4),type = "scatter",mode = 'lines'
            ,name = paste("Mean Salary in USD =",mu))

```

```{r echo=TRUE}
## Systematic Sampling(Unequal Probabilities)
# UPsystematic

set.seed(6000)
#sample size = 50
n=100

#Calculate the inclusion probabilities using the Earnings variable
pik <- inclusionprobabilities(dsSal$salary_in_usd, n)
s <- UPsystematic(pik)

#getting sample from systematic sampling with unequal probabilities
sample.2 <- dsSal[s != 0,]
x2<- sample.2$salary_in_usd
m = median(x2)
mu = round(mean(x2),2)
sd = sd(x2)

heading = "Systematic Sampling"
#pTitle= list(text=heading, y = 0.98, x = 0.5, xanchor = 'center', yanchor =  'top')
pTitle= list(text=heading)
xTitle = list(title = "Salary in USD")
yTitle = list(title = "Probability")
p2=plot_ly(x = ~x2,type = "histogram",name = "Systematic Sampling(Unequal Proabilities)",
         histnorm = "probability",nbinsx = 18)  %>%
layout(title = pTitle,
       xaxis = xTitle, 
       yaxis = yTitle,
       legend = list(x = .6,y=.9)) %>% 
  # add_trace(x = c(m,m),y = c(0,0.22),type = "scatter",mode = 'lines'
  #           ,width =2,name = paste("Median Salary in USD =",m) )|>
  add_trace(x = c(mu,mu),y = c(0,0.22),type = "scatter",mode = 'lines'
            ,name = paste("Mean Salary in USD =",mu))


```

```{r echo=TRUE}
## Systematic Sampling (Equal Probabilities)

#### 3.5. Example – Systematic Sampling

set.seed(6000)
#

N <- nrow(dsSal)
n <- 100
k <- ceiling(N / n)
r <- sample(k, 1)

# select every kth item

s <- seq(r, by = k, length = n)

sample.4 <- dsSal[s, ]
x4 <- sample.4$salary_in_usd
x4 <- x4[!is.na(x4)]
m = median(x4)
mu = round(mean(x4),2)
sd = sd(x4)

#pTitle= list(text=heading, y = 0.98, x = 0.5, xanchor = 'center', yanchor =  'top')
pTitle= heading
xTitle = list(title = "Salary in USD")
yTitle = list(title = "Probability")
p4=plot_ly(x = ~x4,type = "histogram",name = "Systematic Sampling(Equal Probabilities)",
         histnorm = "probability",nbinsx = 18)  %>%
       add_trace(x = c(mu,mu),y = c(0,0.22),type = "scatter",mode = 'lines'
            ,name = paste("Mean Salary in USD =",mu))
  

```

```{r echo = TRUE}
#stratified sample using proportional sizes based on the company size variable


set.seed(6000)
#getting the proportion by company_size
# Proportion

freq <- table(dsSal$company_size)
#getting the sample size of each department for n = 100
sizes <- round(n * freq / sum(freq))
sizes[3] = 1


# getting the stratified sample data for size = n = 100
st <-strata(dsSal, stratanames = c("company_size"),
                       size = sizes, method = "srswor")
sample.3 = getdata(dsSal,st)

x3 = sample.3$salary_in_usd
m = median(x3)
mu = round(mean(x3),2)
sd = sd(x3)

heading = "Stratified Sampling on Company_Size"
#pTitle= list(text=heading, y = 0.98, x = 0.5, xanchor = 'center', yanchor =  'top')
pTitle= list(text=heading)
xTitle = list(title = "Salary in USD")
yTitle = list(title = "Probability")
p3=plot_ly(x = ~x3,type = "histogram",name = "Stratified Sampling",
         histnorm = "probability",nbinsx = 18)  %>%
layout(title = pTitle,
       xaxis = xTitle,
       yaxis = yTitle,
       legend = list(x = .6,y=.9)) %>% 
  add_trace(x = c(mu,mu),y = c(0,0.22),type = "scatter",mode = 'lines'
        ,name = paste("Mean Salary in USD =",mu))


subplot(p,p1,p2,p4,p3,shareX = TRUE,nrows =5) %>% 
  layout( title = "Population vs Sampling Distrubutions",
          xaxis =list(title = " Salary in USD",
                       nticks = 20 ),
          yaxis = list(title = "Probability"))


```

## Findings and Inference

1.  By looking visually at the histograms, as well as the actual values of the
    calculated means (as shown in the legend), the mean for most sampling
    methods are close to the population mean. The exception is Systematic
    Sampling (Unequal Probabilities).
2.  Since Systematic Sampling (Unequal Probabilities) gave more weight to the
    higher salaries, the mean of \$179059.41 for this method is higher compared
    to the population mean of \$157799.72.
3.  The mean \$159979.29 of the Stratified sampling method is closest to the
    population mean of \$157799.72. Overall this is the best sampling method for
    estimating the population mean.
4.  Simple Random Sampling and Systematic Sampling(Equal Probabilities) have
    means of \$160799.72 and \$161811.01.

# Job Counts by Company Size

```{r echo = TRUE}
h = "AI ML Data Science Joby by Company Size<br>(S: Small,M: Medium,L: Large)"
fig = plot_ly(count(dsSal,company_size),type = "pie", labels= ~company_size,
          values = ~n,textposition = 'inside', textinfo = 'label+percent',
          width = 400, height = 400) %>% 
        layout(title = list(text= h,font = list(size =14,color ="darkblue"
                                                      ,x=0.3)))
        
fig

```

## Findings and Inference

About 94.8 % of the jobs in the data sciences sector are in medium companies(50
to 250 employees) and 4.2% in the large companies(\> 250 employees). The reason
for this could be that more employees from medium size companies took this
survey compared to the large and small size companies. Hence we may need further
study on this.

# Analysis of Top 10 Job Titles (By Frequency)

## Top 10 Job Titles(By Frequency)

```{r echo=TRUE}
#Top 10 job titles by Count AI ML Data Science Jobs

data= data.frame(sort(table(dsSal$job_title),decreasing = TRUE))

top10JobTitles <-data[1:10,]  # 99 % jobs in top 25 countries

pTitle = list(text="Top 10 Common Job Titles", y = 0.98,
               x = 0.5, xanchor = 'center', yanchor =  'top')
p5 <- plot_ly(top10JobTitles)%>% 
  add_trace(y = ~Freq, 
            x = ~Var1,
            type ="bar",
            text = ~Freq,
            textposition = "inside",
            marker = list(color = c("cyan",
                                    "yellow",
                                    "orange",
                                    "blue",
                                    "silver",
                                    "green",
                                    "pink"),
                   opacity = rep(0.7, 7))) %>% 
  layout(title = pTitle,
         xaxis = list(title ="",zeroline = FALSE),
         yaxis = list(title = "Count",
                      zeroline = FALSE))
p5
```

### Findings and Inference

The top 4 job titles for AI ML Data Science Jobs in the data set are Data
Scientist (4506) followed by Data Engineer(3963), Data Analyst(2715), Machine
Learning Engineer(2140). These outnumber the other six job titles of Research
Scientist(880), Applied Scientist(600),Data Architect(501), Research
Engineer(451) and Business Intelligence Engineer(296).

## Salary Distribution of Top 10 Job Titles

What are the average salaries by top 10 job titles and experience level

```{r echo = TRUE}

bdata =  filter(dsSal, job_title %in% top10JobTitles$Var1) 
 
h = "Average Salaries by Top 10 Job Titles and Experience"

plot_ly(bdata,x= ~salary_in_usd, color=~job_title,type = "box") %>%
  layout(title = list(text ="Salary Distribution of Top 10 Job Titles",
         font = list(color = "darkblue")),
         xaxis = list(title = "Salary in USD",nticks = 20))
```

### Findings and Inference

Among the top 10 most common job titles in the sector:

1.  The top 4 highest paid jobs are Research Scientist, Research Engineer,
    Applied Scientist, Machine Learning Engineer. They have the highest inter
    quartile range, highest median and the highest a upper fence of \$350,000 to
    \$400,000. These job titles all have fewer than 10 outliers (right side)
    except Machine Learning Engineer, which has more.
2.  Though the frequency of Research Scientist, Research Engineer, Applied
    Scientist are in the bottom ten (of the top 10 most common job titles), they
    are highly paid jobs. This could be because they may require a higher skill
    set with higher educational qualification, indicated by the words ‘Research’
    and ‘Scientist’ in the job title.
3.  Among the top four highest paid job titles, Machine Learning Engineer is the
    only one which is in the top 4 most common job titles as well.
4.  The next three highest paid jobs are Data Scientist, Data Engineer and Data
    Architect Jobs with upper fence salary near \$300,000 to \$350,000. The Data
    Scientist and Data Architect have a lot of outliers in the range of
    \$350,000 to \$400,000. This indicates that these jobs are in high demand
    and the person with the right skills, qualifications, and experience can get
    a salary greater than the market rate.
5.  The bottom 3 job titles in terms of salary are Data Analyst, Analytics
    Engineer and Business Intelligence Engineer. All these have a upper fence of
    \$200,000 to \$280,000.

## Average Salary Trends of Top 10 Job Titles by Experience

What are the average salaries by top 10 job titles and experience level

```{r echo = TRUE}

jdata =  filter(dsSal, job_title %in% top10JobTitles$Var1) |>
 group_by(job_title,experience_level) |>
  summarize(count = n(),avg = mean(salary_in_usd)) |>
  arrange(desc(count))
h = "Average Salaries by Top 10 Job Titles and Experience"
h = paste(h,"<br>(EN: Entry,EX: Executive,MI: Mid Level,SE: Senior)")
plot_ly(jdata,type = "bar", x = ~job_title,y=~avg, color=~experience_level
        )  %>%
  layout(title =h)
  



```

### Findings and Inference

1.  The Data Analyst job has the lowest average for across all years of
    experience compared to the other jobs,
2.  As expected people with Experience level = EX(Executive) have the highest
    average salaries across all job titles, except Research Scientist. For
    Research Scientist job title the people with Experience level = SE(Senior)
    have a slightly higher average salary than with Experience level =
    EX(Executive). The average salary for Execuitve level is highest for Machine
    Learning Engineer followed by Data Scientist.
3.  For entry level jobs Experience level = EN(entry), the highest average
    salary is for the Research Scientist and Research Engineer jobs. The lowest
    entry level salary is for Data Analyst, Data Architect and Business Level
    Engineer jobs.
4.  The same trend as that of Experience level = EN(entry) holds true for
    experience levels MI: Mid level and SE: Senior.

## Average Salary of Top 10 Job Titles by Year

```{r echo = TRUE}

dsSal$work_year = as.character(dsSal$work_year)
kdata =  filter(dsSal, job_title %in% top10JobTitles$Var1) |>
 group_by( job_title, work_year) |>
  summarize(count = n(),avg = round(mean(salary_in_usd))) |>
  arrange(desc(count))


h = "Average Salary By Job Title and Year<br>(For Top 10 Common Job Titles)"

plot_ly(kdata,type = "bar", x = ~job_title,y=~avg, color=~work_year,
        text = ~avg,textposition = "top",textangle =0) %>%
  layout(title = h,xaxis = list(title = ""),
         yaxis = list(title = "Average Salary(USD)"))
  


```

### Findings and Inference

1.  Data Analyst, Data Engineer and Analytics Engineer job titles show the most
    percent increase in average salaries across the years. Of these Data Analyst
    job has had an increase almost every year in average salary. These roles are
    among the lowest paid in the 10 top most common jobs,
2.  For the higher paying jobs of average salaries have held more or less
    steady.
3.  For the Research scientist an abnormally high average salary in 2020
    indicates the presence of an outlier.

# Top 10 Job Titles with Highest Average Salary

Analyze trends for the top 10 job titles with the highest average salary by
experience level.

```{r echo = TRUE}
data10 =   group_by(dsSal,job_title, experience_level) |>
  summarize(count = n(),avg = mean(salary_in_usd)) |>
  arrange(desc(avg))
data10 = data10[1:10,]
xdata = filter(dsSal,job_title %in% data10$job_title)

plot_ly(xdata,x= ~salary_in_usd, color=~job_title,type = "box") %>%
  layout(title = list(text ="Salary Distribution: Top 10 Job Titles by Highest Average Salary",
         font = list(color = "darkblue",width = 500,hright = 500)),
         xaxis = list(title = "Salary in USD",nticks = 20))
h = "Count of Top 10 Job Titles by Experience Level with Highest Average Salary "
h = paste(h,"<br>(EN: Entry,EX: Executive,MI: Mid Level,SE: Senior Level)")
plot_ly(data10,y = ~count, 
            x = ~job_title,
            type ="bar",
            text=~count, 
            textposition = "inside",
            color = ~experience_level) %>% 
  layout(title = h,yaxis = list(title = "Count"))
        



```

## Findings and Inference

1.  By looking at the distribution of salaries for the 10 job titles with the
    highest average salary we can see that Prompt Engineer made it to this list
    as they have one or two outliers with a very high salary. These jobs would
    not be in the list if it were not for these outliers.

2.  Jobs of AWS Data Architect, Finance Data Analyst, Cloud Data Architect and
    Data Science Tech lead only have 1 record in the data set and hence are
    outliers.

3.  The top paid jobs of Machine Learning Developer, Deep Learning Engineer,
    Applied Data Scientist, Head of Machine Learning all are held by people
    whose experience level is Senior or Executive level.

4.  The counts for all of the 10 job titles with highest average salary is less
    than 10 records.

# Job Trends By Experience Level

What is the proportion of jobs in the market by experience level.

```{r echo = TRUE}
h = "Job Count by Experience Level"
h = paste(h,"<br>(EN: Entry,EX: Executive,MI: Mid Level,SE: Senior)")
fig = plot_ly(count(dsSal,experience_level),type = "pie", 
              labels= ~experience_level,values = ~n,
              textposition = 'inside', textinfo = 'label+percent',
              width = 400, height = 400) %>% 
        layout(title = list(text= h,font = list(size =14,color ="darkblue"
                                                      ,x=0.3)))
        
fig
```

## Findings and Inference

Though one would expect entry level jobs to be the most frequent in a job
market, it looks like Senior Level Jobs are the most frequent at 64.5 % in the
AI ML Data Science sector. This is followed by the mid level experience jobs at
25.2%, and the entry level jobs at 7.6%. The executive level jobs are very few
at 2.8%.

Since the number of entry level jobs has the lowest proportion, it may be
difficult for a person to get an entry level job in this sector. Given the
higher proportion of senior and mid level jobs, higher educational
qualification, more work experience or higher skills may be required to get a
job in this sector.

# Salary Trends By Experience

Are the salaries higher for people with more years of work experience.

```{r echo = TRUE}
h = "Salary Distribution<br>(EN: Entry,EX: Executive,MI: Mid Level,SE: Senior)"
plot_ly(dsSal,x = ~salary_in_usd, color=~experience_level,type = "box") %>%
  layout(title = h,
         xaxis = list(title = "Salary in USD",nticks = 20))


```

## Findings and Inference

We can see that the salaries in this industry are definitely higher for people
with more job experience. People at the Executive level are paid the highest
followed by people at Senior level. As expected entry level jobs are paid the
lowest.

All 4 experience levels have outliers, with the senior level having the most,
followed by the mid level, executive level and entry level.

# Average Salary Trends by Year

Trends in Average Salary from 2020 to 2024. Has the average salary in this
sector increased during this period, decreased or remained the same. This is
compared with the count of survey respondents by year to analyze the data
better.

```{r echo = TRUE}

dsSal$work_year = as.numeric(dsSal$work_year )
top5JobTitles = top10JobTitles[1:5,]
data5 =  filter(dsSal,job_title %in% top5JobTitles$Var1) |>
  group_by(work_year) |>
  summarize(count = n(),avg = round(mean(salary_in_usd)))|>
  arrange(work_year)


h = "Average Salary and Count of Survey Respondents by Work Year"

subplot(

  plot_ly(data5,type = "scatter", mode = "lines",x = ~work_year,y=~avg
        ,name = "Average Salary")  %>%
 layout(title =h,xaxis = list(title = "Work Year"),
        yaxis = list(title = "Average Salary"),
        xaxis =list(range = list(2019, 2025))),
  plot_ly(data5,type = "scatter", mode = "lines",x = ~work_year,y=~count,
            name = "Count")  %>%
     layout(title =h,xaxis = list(title = "Work Year"),
        yaxis = list(title = "Count of Respondents"),
        xaxis =list(range = list(2019, 2025))),
 nrows = 2 ,titleY = TRUE,shareX = TRUE,heights = c(0.55,0.45)) |> 
  layout(xaxis = list(title = "Work Year",range = list(2019, 2024)))
```

## Findings and Inference

We see that the average salary in the AI ML Data Science job sector had a linear
increase from \$123372(in 2021) to \$159822(in 2023), and plateaued \$159522(in
2024). There is a sudden drop in average salary from \$163284(in 2020) to
\$123372(in 2021). This seems to be an anomaly and could be because the number
of survey respondents was low at 23 in 2020. There is a yearly increase in
number of survey respondents in the US from 2020 to 2024. This could be because
the number of data science jobs has seen an increase in the past few years due
to the boom in AI and ML.

# Trends in Remote Work

Trends in Remote Work from 2020 to 2024. How has the remote work scene shifted
in the time period.

```{r echo = TRUE}

dsSal$work_year = as.character(dsSal$work_year)
dsSal$remote_ratio = as.character(dsSal$remote_ratio)
mdata = group_by(dsSal, work_year,remote_ratio) |>
  summarize(count = n()) |>
  arrange(desc(work_year))


h = "Remote Work Ratio by Work Year"
h = paste(h,"<br>(0: No Remote Work,100: Fully Remote,50: Hybrid)")
plot_ly(mdata,type = "bar", x = ~work_year,y=~count, color=~remote_ratio
        )  %>%
 layout(title =h,xaxis = list(title = "Work Year"),
        yaxis = list(title = "Count"))


```

## Findings and Inference

1.  We can ignore the data for 2020 and 2021 as very few people have filled out
    the survey in those years.
2.  We can see that the number remote jobs have decreased compared to 'no remote
    work' jobs from 2022 to 2024.
3.  In 2024 the 'no remote work' jobs are almost 4 times that of remote jobs,
    indicating an increased trend of employers wanting the employees to work in
    person, rather than from home.
4.  The number of people working hybrid jobs is low compared to remote and
    in-person jobs. This could be because people working hybrid (more than 50%
    in person work) may have identified themselves as 'no remote work' in the
    survey.

# Conclusion

Though the data set provides good insight into the various analyses and trends
of the AI ML Data Sciences job market in the US, it is still based on surveys
from a website and may not be representative of the whole population. Hence this
study may be used as a start and further studies conducted to come to an
unbiased conclusion. This survey could also benefit from having a question on
the educational qualification and the major of the employees to provide more
insight. In spite of the above limitations the data set does a give a good idea
of the AI ML Data Sciences sector for the US job market and is a good starting
point for future research. This study is also relevant given the boom in the AI
ML Data Science job sector over the past few years.
